Multidimensional Persistence Module Classification via Lattice-Theoretic Convolutions

### 2020-11-19 paper/paper.tex
minor edits to paper from reviewer comment

% to avoid loading the natbib package, add option nonatbib:
\usepackage[preprint]{neurips_2020_tda}

@@

\author{%
  Hans Riess  \\
  Department of Electrical and Systems Engineering \\
  University of Pennsylvania\\
  Philadelphia, PA 19104 \\
  \texttt{hmr@seas.upenn.edu} \\
  % examples of more authors
  \And
  Jakob Hansen \\
  Department of Mathematics \\
  Ohio State University \\
  Columbus, OH 43210 \\
  \texttt{hansen.612@osu.edu}
  % Address \\
  % \texttt{email} \\
  % \AND
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
  % \And
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
  % \And
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
}

@@

Multidimensional persistence generalizes single-dimensional persistent homology
in order to tackle filtrations parameterized in multiple dimensions.
% Such
% filtrations arise, for example, in a Rips filtration of a finite metric space
% for which the finite metric space itself carries a filter function. %is parameterized by a signal
% %attributing a vector in $\R^d$ to every point of the metric space.
% Restricting
% to points in the metric space appearing below a certain coordinate-wise
% threshold results in a multi-filtration of the corresponding Rips complexes.
%Multidimensional persistence boasts stability properties similar to those in one
%dimension~\cite{lesnick_theory_2015}.
Unfortunately, there is no complete
compact barcode-like characterization of multidimensional persistence
modules~\cite{carlsson_theory_2009}. We must make do with incomplete invariants.
There are various algebraic invariants; in this paper we will use the
\textit{Hilbert function} and the \textit{multi-graded Betti numbers}\footnote{\textit{Caveat lector:} the multi-graded (algebraic) Betti numbers need not be confused with the topological Betti numbers, i.e.~the rank of homology.}, both of
which are $\mathbb{N}$-valued functions on the parameter space.
%Their names arise from their relationship with the homological algebra of
%persistence modules.
The Hilbert function is nothing more than the pointwise
(topological) Betti numbers, and the multi-graded Betti
numbers have a geometric interpretation in terms of births and deaths~\cite{knudson_refinement_2008}.

% The \textit{rank
% invariant}, another invariant, is shown to be complete in the case of single
% dimensional persistence \cite{carlsson_theory_2009}; due to its difficulty to
% compute it is not considered here.

Multidimensional persistence suffers two hindrances to its usefulness in
machine learning. First, software for computing multidimensional persistence is
scarce. To the authors' knowledge, RIVET is the only available software for
computing multidimensional persistence~\cite{lesnick_interactive_2015}; RIVET
specializes to $2$-dimensional persistence and is focused on interactive
visualization rather than machine-interpretable output.
% provides a user-interface for
% visualizing multi-filtrations as well as a pipline for computing the Hilbert
% function and the multi-graded Betti numbers in degrees $0$, $1$, and $2$.
Second, %for all the attention multidimensional persistence has received,
there has been no activity, as far as the authors are aware, studying the extraction
of suitable features for multidimensional persistent homology.
%Carlsson has a little bit in a recent paper...

We hope to ignite an interest in filling both of these gaps. In this paper, we
propose a naive featurization of multidimensional persistence modules based on
the aforementioned invariants, and design an architecture for classifying these
persistence modules. This architecture employs a lattice-theoretic notion of
convolution, thereby respecting the order relation of the parameters of the
persistence module. We implement our model and compare the performance of our
proposed lattice-convolutional architecture with a (simplified) standard
convolutional architecture.

@@

Applying the simplicial homology functor (with coefficients in a field $k$)
$H_i$ to $\text{Rips}_r(\mathcal M)$ produces a sequence of vector spaces
$PH_i(r)$. The inclusions $\text{Rips}_r(\mathcal M) \to
\text{Rips}_{r'}(\mathcal M)$ induce maps $PH_i(r) \to PH_i(r')$, producing the
data of a \introduce{persistence module}. This structure can be compactly
described as a functor from $\R$, viewed as a category via its standard order
structure, to the category $\Vect_k$ of vector spaces over $k$. The simplicity
of the category $\R$ gives these persistence modules simple structure: they
decompose as direct sums of interval modules $I_{[a,b)}$, which have
$I_{[a,b)}(r) = k$ for $a \leq r < b$ and zero otherwise. The maps are the
identity where possible and the zero map otherwise.
% is this the right shape for the intervals?

% This representation theoretic fact gives a representation\footnote{Pun intended.} of
% $\R$-indexed persistence modules via \introduce{barcodes} or
% \introduce{persistence diagrams}. Each bar $I_{[a,b)}$ in the barcode represents
% a homology class which is \introduce{born} at $a$ and \introduce{dies} at $b$.

@@

While there does not exist a discrete set of invariants for  $PH_i$, we can extract
meaningful features. Two particularly informative types of features are 
 the \introduce{Hilbert function}
  \[\text{Hilb}: \R_+ \times \R \to \Z_+;\quad \mathbf{x} \mapsto \dim(M_\mathbf{x}),\] and
the \introduce{multi-graded Betti numbers} $\xi_j: L \to \Z_+$, for $j =
0,1,2$. For $PH_i$ as above, the Hilbert function counts the number of
connected components ($i=0$), cycles ($i=1$), or higher dimensional voids ($i >
1$) of the complex $\mathbb{X}_{r,t}$ at each $(r,t) \in \R_+\times \R$. The multi-graded
Betti numbers, on the other hand, capture information about births
and deaths of homology classes.

@@

Convolutional neural networks are uesful in part because the convolution kernels
(here the functions $g$) can have very small support, reducing the number of
parameters that must be learned. In the standard convolutional setting, these
kernels are implicitly supported in a neighborhood of the origin, but the
location of the kernel is not usually explicitly specified. In the lattice
setting, we do need to specify where the kernel resides. In the abelian group
case, the kernel is supported near the identity, and similarly, 
when we treat our domain as a lattice, the kernel should be supported
near the neutral element of the operation. That is, for a meet convolution, $g$
should have support near the maximum $(m,n)$, and for a join convolution, $g$
should have support near the minimum $(0,0)$.

@@

The pipline thus begins with a 3D polyhedral model, of which $3000$ vertices are sampled to produce a point
cloud in $\R^3$. This point cloud then produces a bifiltered simplicial complex,
whose persistent homology we calculate using
RIVET~\cite{lesnick_interactive_2015}, sampled at a discrete grid of $40 \times
40$ points, producing lattice-indexed signals given by the Hilbert function and
the multi-graded Betti numbers $\xi_0, \xi_1, \xi_2$; four features in total. These are then passed to the classifier, which
produces a class prediction, in this case either ``sofa'' or ``monitor.''

@@

We compare the performance of two convolutional networks on this classification
task. One uses the lattice-convolution based layers described in
Section~\ref{sec:latticeCNN}, and the other uses standard convolutional layers.
Each has three convolutional layers followed by three fully connected layers.
The lattice-based convolution layers are of the form $\text{MeetConv}(x) +
\text{JoinConv}(x)$. All convolution kernels have dimension $4 \times 4$, hidden
convolution layers have 16 features, and the final convolution has 8 features.
The fully connected layers have 32 features at each inner layer.
We use a cross entropy loss function with a softmax in the final layer.
The networks are trained using the Adam gradient algorithm with learning rate
$0.001$ for 50 epochs. We reserve 10\% of the data for testing. Results are
shown in Figure~\ref{fig:comparison}.

The lattice-based convolution and classical convolution networks perform
similarly on the training set. However, the lattice convolution network's
predictions generalize more accurately to the testing set ($\approx 7\%$ improvement in testing accuracy).
This improvement is  at the expense of more variability in performance during
training. (Our code is available at \texttt{https://github.com/alfred-tarski/latticeconv} .)

\begin{figure}[h]
  \caption{Comparison training curves for the lattice neural network and
    standard convolutional neural network. Results are averaged over 5 runs.}\label{fig:comparison}
    \begin{center}
%    \includegraphics[width=0.49\textwidth]{testing-accuracy.png}
%    \vspace{0.2cm}
%    \includegraphics[width=0.49\textwidth]{training-accuracy.png}
	\includegraphics[width=0.85\textwidth]{figure.png}
    \end{center}
\end{figure}
\section{Discussion}
Our proposed featurization for persistence modules is rather naive, but well
adapted to the use of lattice convolutions as a data processing method. The
lattice convolutional neural network shows promise as a method for classifying
features arising from a multiparameter persistence module. The algebraic
perspective on partially ordered sets, exemplified by lattices, may also offer
approaches to featurizing more complex invariants of persistence modules. In
particular, the incidence algebra may offer a natural way to represent the rank
invariant~\cite{carlsson_theory_2009} in a way amenable to convolution-like
operations.
We hope with these brief experiments to inspire further work on featurizing
multidimensional persistence for use in machine learning algorithms.
% anything else?

>>>>

% to avoid loading the natbib package, add option nonatbib:
\usepackage[final]{neurips_2020_tda}

@@

\author{%
  Hans Riess \\
  University of Pennsylvania\\
  \texttt{hmr@seas.upenn.edu} \\
  % examples of more authors
  \And
  Jakob Hansen \\
  Ohio State University \\
  \texttt{hansen.612@osu.edu}
  % Address \\
  % \texttt{email} \\
  % \AND
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
  % \And
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
  % \And
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
}

@@

Multidimensional persistence generalizes single-dimensional persistent homology
in order to tackle filtrations parameterized in multiple dimensions.
% Such
% filtrations arise, for example, in a Rips filtration of a finite metric space
% for which the finite metric space itself carries a filter function. %is parameterized by a signal
% %attributing a vector in $\R^d$ to every point of the metric space.
% Restricting
% to points in the metric space appearing below a certain coordinate-wise
% threshold results in a multi-filtration of the corresponding Rips complexes.
%Multidimensional persistence boasts stability properties similar to those in one
%dimension~\cite{lesnick_theory_2015}.
Unfortunately, there is no complete
compact barcode-like characterization of multidimensional persistence
modules~\cite{carlsson_theory_2009}. We must make do with incomplete invariants.
There are various algebraic invariants; in this paper we will use the
\textit{Hilbert function} and the \textit{multi-graded Betti numbers}\footnote{\textit{Caveat lector:} the multi-graded (algebraic) Betti numbers need not be confused with the topological Betti numbers, i.e.~the rank of homology.}, both of
which are $\mathbb{N}$-valued functions on the parameter space.
%Their names arise from their relationship with the homological algebra of
%persistence modules.
The Hilbert function is nothing more than the pointwise
(topological) Betti numbers, and the multi-graded Betti
numbers have a geometric interpretation in terms of births and deaths~\cite{knudson_refinement_2008}.
 The \textit{rank
 invariant}, another invariant, is shown to be complete in the case of single
 dimensional persistence \cite{carlsson_theory_2009}; due to its difficulty to
 compute it is not considered here.

Multidimensional persistence suffers two hindrances to its usefulness in
machine learning. First, software for computing multidimensional persistence is
scarce. To the authors' knowledge, RIVET is the only available software for
computing multidimensional persistence~\cite{lesnick_interactive_2015}; RIVET
specializes to $2$-dimensional persistence and is focused on interactive
visualization rather than machine-interpretable output.
% provides a user-interface for
% visualizing multi-filtrations as well as a pipline for computing the Hilbert
% function and the multi-graded Betti numbers in degrees $0$, $1$, and $2$.
Second, %for all the attention multidimensional persistence has received,
there has been only very recent and preliminary activity \cite{vipond2020multiparameter,carriere2020multiparameter} focusing on extraction
of suitable features for multidimensional persistent homology. A full-fledged deep learning pipeline
for multidimensional persistence waits in the wings.
%Carlsson has a little bit in a recent paper...

We hope to ignite further interest in filling both of these gaps. In this paper, we
propose a naive featurization of multidimensional persistence modules based on
the aforementioned invariants, and design an architecture for classifying these
persistence modules. This architecture employs a lattice-theoretic notion of
convolution, thereby respecting the order relation of the parameters of the
persistence module. We implement our model and compare the performance of our
proposed lattice-convolutional architecture with a (simplified) standard
convolutional architecture.

\subsection*{Related work}

Very recent advances have been made to featurize multidimensional persistence modules via
landscapes \cite{vipond2020multiparameter} and images \cite{carriere2020multiparameter}.
Our naive approach at featurizing $2$-dimensional persistence modules has the advantage
of being readily computable with existing software \cite{lesnick_interactive_2015}.
Persistence modules supported on lattices are of increased interest.
By leveraging M\"{o}bius inversion, these persistence modules are shown to generate a stable persistence diagram \cite{mccleary2020edit} as well as factor through a nice category with pseudo-inverses \cite{krishnan2020invertibility}.

@@

Applying the simplicial homology functor (with coefficients in a field $k$)
$H_i$ to $\text{Rips}_r(\mathcal M)$ produces a sequence of vector spaces
$PH_i(r)$. The inclusions $\text{Rips}_r(\mathcal M) \to
\text{Rips}_{r'}(\mathcal M)$ induce maps $PH_i(r) \to PH_i(r')$, producing the
data of a \introduce{persistence module}. This structure can be compactly
described as a functor from $\R$, viewed as a category via its standard order
structure, to the category $\Vect_k$ of vector spaces over $k$. The simplicity
of the category $\R$ gives these persistence modules simple structure: they
decompose as direct sums of interval modules $I_{[a,b)}$, which have
$I_{[a,b)}(r) = k$ for $a \leq r < b$ and zero otherwise. The maps are the
identity where possible and the zero map otherwise.

%This representation theoretic fact gives a representation (\textit{no pun intended}) of
%$\R$-indexed persistence modules via \introduce{barcodes} or
%\introduce{persistence diagrams}. Each bar $I_{[a,b)}$ in the barcode represents
%a homology class which is \introduce{born} at $a$ and \introduce{dies} at $b$.

@@

While there does not exist a complete discrete invariant for  $PH_i$, we can extract
meaningful features. Two particularly informative types of features are 
 the \introduce{Hilbert function}
  \[\text{Hilb}: \R_+ \times \R \to \Z_+;\quad (r,t) \mapsto \dim(PH_i(r,t)),\] and
the \introduce{multi-graded Betti numbers} $\xi_j: \R_+ \times \R \to \Z_+$, for $j =
0,1,2$. For $PH_i$ as above, the Hilbert function counts the number of
connected components ($i=0$), cycles ($i=1$), or higher dimensional voids ($i >
1$) of the complex $\mathbb{X}_{r,t}$ at each $(r,t) \in \R_+\times \R$. The multi-graded
Betti numbers, on the other hand, capture information about births
and deaths of homology classes.\footnote{ $\xi_j(r,t)$ counts the rank of the $j$th term of the free resolution of $PH_i$ at grading $(r,t)$ \cite{lesnick_interactive_2015}.}

@@

Convolutional neural networks are useful in part because the convolution kernels
(here the functions $g$) can have very small support, reducing the number of
parameters that must be learned. In the standard convolutional setting, these
kernels are implicitly supported in a neighborhood of the origin, but the
location of the kernel is not usually explicitly specified. In the lattice
setting, we do need to specify where the kernel resides. In the abelian group
case, the kernel is supported near the identity, and similarly, 
when we treat our domain as a lattice, the kernel should be supported
near the neutral element of the operation. That is, for a meet convolution, $g$
should have support near the maximum $(m,n)$, and for a join convolution, $g$
should have support near the minimum $(0,0)$.

@@

The pipline thus begins with a 3D polyhedral model, of which $3000$ vertices are sampled to produce a point
cloud in $\R^3$. This point cloud then produces a bifiltered simplicial complex,
whose persistent homology in degree zero we calculate using
RIVET~\cite{lesnick_interactive_2015}, sampled at a discrete grid of $40 \times
40$ points, producing lattice-indexed signals given by the Hilbert function and
the multi-graded Betti numbers $\xi_0, \xi_1, \xi_2$; four features in total. These are then passed to the classifier, which
produces a class prediction, in this case either ``sofa'' or ``monitor.''

@@

We compare the performance of two convolutional networks on this classification
task. One uses the lattice-convolution based layers described in
Section~\ref{sec:latticeCNN}, and the other uses standard convolutional layers.
Each has three convolutional layers followed by three fully connected layers.
The lattice-based convolution layers are of the form $\text{MeetConv}(x) +
\text{JoinConv}(x)$. All convolution kernels have dimension $4 \times 4$, hidden
convolution layers have 16 features, and the final convolution has 8 features.
The fully connected layers have 32 features at each inner layer.
We use a cross entropy loss function with a softmax in the final layer.
The networks are trained using the Adam gradient algorithm with weight decay with learning rate
$5 \cdot 10^{-4}$ for 50 epochs. We reserve 10\% of the data for testing. Results are
shown in Figure~\ref{fig:comparison}.

The lattice-based convolution and classical convolution networks perform
similarly on the training set. However, the lattice convolution network's
predictions generalize more accurately to the testing set ($\approx 7\%$ improvement in testing accuracy).
This improvement is  at the expense of more variability in performance during
training.
%(Our code is available at \texttt{https://github.com/alfred-tarski/latticeconv}.)

\begin{figure}[h]
  \caption{Comparison training curves for the lattice neural network and
    standard convolutional neural network.}\label{fig:comparison}
    \begin{center}
%    \includegraphics[width=0.49\textwidth]{testing-accuracy.png}
%    \vspace{0.2cm}
%    \includegraphics[width=0.49\textwidth]{training-accuracy.png}
%	\includegraphics[width=0.85\textwidth]{figure.png}
    \end{center}
\end{figure}
\section{Discussion}
Our proposed featurization for persistence modules is rather naive, but well
adapted to the use of lattice convolutions as a data processing method. The
lattice convolutional neural network shows promise as a method for classifying
features arising from a multiparameter persistence module. The algebraic
perspective on partially ordered sets, exemplified by lattices, may also offer
approaches to featurizing more complex invariants of persistence modules. In
particular, the incidence algebra may offer a natural way to represent the rank
invariant~\cite{carlsson_theory_2009} in a way amenable to convolution-like
operations.
We hope with these brief experiments to inspire further work on featurizing
multidimensional persistence for use in machine learning algorithms.
% anything else?
